Identifiyng Inter-Genomic Repeats Using Fast, Approximate Betweenness Centrality

### 2015-05-06 manuscript/manuscript.tex
report modified

Metagenomic data are considerably more complex. The assembly of metagenomic data is complicated due to several reasons. Few of them are: (i) widely different levels of representation for different organisms in a community; (ii) genomic variation between closely related organisms; (iii) conserved genomic regions shared by distantly related organims; and (iv) repetitive sequences within individual genomes. Genomic repeats are the major challenge when assembling isolated genomes and their effect is prominently seen in metagenomic datasets. It has been show by Kingsford et al.\cite{kingsford} that the number of reconstructions grow exponentially with the number of repeats. Therefore it is impossible find one correct solution for assembly. There are several approaches such as \cite{bambus} to find repeats in metagenomic data. The main drawback of these methods is that they do not scale well as number of nodes in assembly graph increase and prove very inefficient as number of nodes reaches to about a million. To tackle this problem, we explore some of the work in social network analysis, particularly in the area of node centrality. We make use of betweenness centrality to find out repeats in assembly graph.

@@

Geisberger et al. \cite{sanders} have shown that $X := 2n\delta(v)$ is an unbiased estimator of centrality of v that is $E(X) := g(v)$. Therefore by averaging $k$ independent runs of this estimator, we get an approximation $\frac{X_{1}+X_{2}+..+X_{k}}{k}$ of the betweenness centrality for node $v$.

To compute the contributions of nodes to centrality, a modified scheme of Brandes algorithm is used. Contributions are calculated as follows:

$$\delta_{s}(v) = \sum_{w \in succ(v)} \frac{\mu(s,v)}{\mu(s,w)}.\frac{\sigma_{sv}}{\sigma_{sw}}(1+\delta_{s}(w))$$
where $\mu(s,w)$ is the shortest path distance from $s$ to $w$. It uses $\frac{\mu(s,v)}{\mu(s,w)}$ in aggregation instead of $1$ in Brandes algorithm.

@@

The Vapnik-Chernovenkis(VC) dimensions of a class of subset is defined on a set of points is a measure of expressiveness of such class. Given a probability distribution on the set of points, a finite bound on the VC-dimension of the class of subsets implies a bound on the number of random samples required to approximate the probability of each subset in the class with its empirical average. This technique is used to derive the expressiveness of the sample of shortest paths derived. For graph $G(V,E)$, let $S_{all}$ be the union of $T_{v}$ for all $v \in V$, where $T_{v}$ is the set of all shortest path to which $v$ is internal. VC dimension technique helps in choosing sample from $S_{all}$ so that it gives probabilistic guarantees. It is show that the upper bound on VC dimensions of $S_{all}$ is given by:
$$VC(S_{all}) \leq \left\lfloor{log_{2}(VD(G) - 2)}\right\rfloor + 1$$
where $VD(G)$ is vertex diameter of $G$. Using this upper bound, sample size $r$ can be estimated as follows:
\begin{equation} \label{sample_size_algo}
r = \frac{c}{\epsilon^{2}}(\left\lfloor{log_{2}(VD(G) - 2)}\right\rfloor + 1 + ln\frac{1}{\delta})
\end{equation} 
where $c$ is a universal constant with value $0.5$. It is guaranteed that betweenness centralities estimated using this sample size are within additive factor $\epsilon$ from real values with probability atleast $1-\delta$.

>>>>

Metagenomic data are considerably more complex. The assembly of metagenomic data is complicated due to several reasons. Few of them are: (i) widely different levels of representation for different organisms in a community; (ii) genomic variation between closely related organisms; (iii) conserved genomic regions shared by distantly related organims; and (iv) repetitive sequences within individual genomes. Genomic repeats are the major challenge when assembling isolated genomes and their effect is prominently seen in metagenomic datasets. It has been shown by Kingsford et al.\cite{kingsford} that the number of reconstructions grow exponentially with the number of repeats. Therefore it is impossible find one correct solution for assembly. There are several approaches such as \cite{bambus} to find repeats in metagenomic data. The main drawback of these methods is that they do not scale well as number of nodes in assembly graph increase and prove very inefficient as number of nodes reaches to about a million. To tackle this problem, we explore some of the work in social network analysis, particularly in the area of node centrality. We make use of betweenness centrality to find out repeats in assembly graph.

@@

This algorithm is efficiently implemented using two techniques. First one is \textit{linear scaling}, where the contribution of a node $v$ depends linearly on the distance to the sample. Nearer the node is to the sample, lesser is its contribution counted. Linear scaling is implemented using a simple variant of aggregation scheme proposed in Brandes algorithm. Second technique is \textit{bisection scaling} where the sample only contributes to the second half of a path. This means that for a vertex $v$, its contribution is considered only if it is at atleast half distance between $s$ and $t$ away from $s$ in forward search or more than half a distance away from $t$ in backward search.

@@

The Vapnik-Chernovenkis(VC) dimensions of a class of subset is defined on a set of points is a measure of expressiveness of such class. Given a probability distribution on the set of points, a finite bound on the VC-dimension of the class of subsets implies a bound on the number of random samples required to approximate the probability of each subset in the class with its empirical average. This technique is used to derive the expressiveness of the sample of shortest paths derived. For graph $G(V,E)$, let $S_{all}$ be the union of $T_{v}$ for all $v \in V$, where $T_{v}$ is the set of all shortest path to which $v$ is internal. VC dimension technique helps in choosing sample from $S_{all}$ so that it gives probabilistic guarantees. It is shown that the upper bound on VC dimensions of $S_{all}$ is given by:
$$VC(S_{all}) \leq \left\lfloor{log_{2}(VD(G) - 2)}\right\rfloor + 1$$
where $VD(G)$ is vertex diameter of $G$. Using this upper bound, sample size $r$ can be estimated as follows:
\begin{equation} \label{sample_size_algo}
r = \frac{c}{\epsilon^{2}}(\left\lfloor{log_{2}(VD(G) - 2)}\right\rfloor + 1 + ln\frac{1}{\delta})
\end{equation} 
where $c$ is a universal constant with value $0.5$. It is guaranteed that betweenness centralities estimated using this sample size are within additive factor $\epsilon$ from real values with probability atleast $1-\delta$.
